Communicative reduction in referring expressions within a multi-player negotation game

### 2023-01-27 write-ups/cogsci23_paper/manuscript.Rmd
edits from Mike; fixed a number of incorrect statements in results

---
title: "Communicative reduction in referring expressions in a multi-player negotation game"
bibliography: [ aaflowers.bib]
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

abstract: >
    Reduction of referring expressions to short conventionalized terms is found robustly in dyadic repeated reference games when the target images are initially difficult to name. Establishing reference is crucial to many conversations with different goals and more complex motivations; thus, it is useful to look at how the findings from reference games, where participants share the explicit goal of establishing joint reference, carry over into situations where participants have different overarching goals. In the current work, we analyse a dataset where reference was embedded in strategic 3- player negotiation and coordination games. In these more complex games, we found that the patterns of reduction and convergence to conventions held across two different incentive conditions, with some suggestive differences between the conditions.

keywords: >
    Convention formation; reference games; reduction of referring expressions; multi-party communication;

@@

Successful communication is grounded in a shared understanding of what utterances mean in the context they are produced in. Whether the communicative goals are cooperative or adversarial, interlocuters need to establish joint reference to be effective. In many cases conventional word meanings are enough, but there are also contexts where objects to be referred to are not easy to distinguish. In these situations, establishing reference is more difficult, but no less important. <!--For one, in everyday communication, establishing reference rarely happens in isolation but is usually a subgoal in a larger conversational goal. The overall goal might be cooperative, such as in asking for a piece of kitchen equipment while cooking together, but it can also be adversarial, such as in a negotiation where the assets to be divied up must be referenced. In the reference game context, the goal is always cooperative, to match the images or get them in the same order. -->

The formation of these ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986; @krauss1964changes]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize with a pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs. This has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, and when they change [@brennan1996conceptual; @metzing2003conceptual; @leung2020parents; @weber2003; @yoonAudienceDesignMultiparty2019].

The implicit theoretical claim is that reduction to partner-specific conventions occurs whenever people interact repeatedly in ways that require reference to some objects without adequate conventional names. Do these phenomena generalize beyond artificial reference game environments?

Our current work begins to address this by looking for these phenomena in a slightly more naturalistic setting. The conversational goals of reference games are not representative of typical language use. In everyday communication, establishing reference is often an instrumental subgoal in a larger conversation. The overall task may be cooperative, such as asking for kitchen equipment while cooking together, or adversarial, such as identifying items while negotiating a division of assets. In the reference game context, the goal is explicitly cooperative: match corresponding images.

```{r interface, fig.env = "figure*", fig.pos = "t!", fig.width=6, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Player interface. Panel A shows the selection phase, where each participant sees 6 flowers, 4 with value bars. Panel B shows the feedback stage, after all players have selected. When multiple players select the same flower, they recieve a lower value rather than what is shown.  \\label{interface}", cache=FALSE}

@@

@mankewitz2021 bridged the gap between reference games with hard-to-name objects and negotation datsets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we separated out the referring expressions in order to check for patterns of convention-formation.

Our current work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge were able to develop conventionalized reference expressions. We explored whether the phenomena found in classic dyadic iterated reference games extended to this more naturalistic domain, and whether they differed across the individual and shared incentive conditions. The key questions we addressed were:

@@

Two games consisting of `r raw_rows` utterances were annotated by both annotators.  `r nrow(both)` utterances were identified as containing reference expressions by at least one annotator; of these annotators agreed on the exact reference expression for `r agree_span` (`r round(agree_span/nrow(both)*100)`%) of the cases. The second annotator only coded the target of the referring expression for `r nrow(count)` of the utterances. Of these, the two annotators agreed on the target in `r agree_target` (`r round(agree_target/nrow(count)*100)`%) of cases. We take this level of interannotator agreement as an indication that the reference spans and targets were identified reliably.

We extracted a total of `r annotated |> filter(numword>0) |>  tally()` referring expressions.

@@

For descriptions of different flowers, cosine similarity decreased over time (trial: `r stats_text(wi_g_bet_f,6)`, see Figure \ref{during} lower left and middle panels). There was not an effect of utterances being produced by the same person (`r stats_text(wi_g_bet_f,8)`), or interactions between who said it and trial number (`r stats_text(wi_g_bet_f,7)`) .

These results are consistent with the reference game phenomenon.

@@

knitr::kable(text_examples, caption = "Examples of descriptions of different flowers from different games, illustrating reduction and convergence phenomenon. Games are numbered, letters refer to players within games. Images of flowers can be found in Figure 1: flower 1 upper left, flower 2 lower center, flower 3 lower left. \\label{extable}",  format="latex", booktabs=TRUE, linesep="") %>%
kable_styling(full_width = F) %>%
  row_spec(c(3,9), hline_after=T) |> 
column_spec(4, width = "12em") %>%
  kable_styling(latex_options="H")

@@

For descriptions of the same flower across games, we found that cosine similarities decreased slightly over time  (trial: `r stats_text(bet_g_wi_f,4)`, see Figure \ref{during} upper right panel). This is weakly consistent with the predictions.

@@

Referring descriptions later in a game were more similar to the convention than descriptions earlier in the game  (trial: `r stats_text(to_e_wi_f,6)`, see Figure \ref{toend} upper panels). Utterances were more similar to the convention given by the same person (versus the convention given by a groupmate) (`r stats_text(to_e_wi_f,8)`), although this did not interact with trial number (`r stats_text(to_e_wi_f,7)`). This is consistent with the expected pattern of convention formation.

@@

### Within games
Games in the shared utilities condition had slightly higher cosine similarities between descriptions of the same flower within games (`r stats_text(wi_g_wi_f,2)`), but this did not interact with trial number (`r stats_text(wi_g_wi_f,3)`) or whether the utterances were produced by the same participant (`r stats_text(wi_g_wi_f,5)`). As a complement to this, games in the shared utilities condition had slightly lower similarities between descriptions of different flowers within games, but this did not interact with trial (`r stats_text(wi_g_bet_f,3)`) or whether the utterances were produced by the same player (`r stats_text(wi_g_bet_f,5)`).

Comparing to the end conventions, games in the shared utility condition produced flower descriptions that were more similar to the convention for that flower (`r stats_text(to_e_wi_f,2)`), but this did not interact with trial number (`r stats_text(to_e_wi_f,3)`) and did not significantly interact with same participant versus different participant (`r stats_text(to_e_wi_f,5)`).  In the shared utility condition, descriptions were slightly less similar to the conventions for other flowers (`r stats_text(to_e_bet_f,2)`), but this did not interact with trial (`r stats_text(to_e_bet_f,3)`). There was an interaction between shared utility condition and the referring expression coming from the same person who gave the convention for the different flower (`r stats_text(to_e_bet_f,5)`). Taken together, groups in the shared utility had more internal alignment on the mapping between flowers and names.

### Between games
Across different games, games in the shared utility condition tended to have more similar descriptions for the same flower (`r stats_text(bet_g_wi_f,2)`); this did not interact with trial (`r stats_text(bet_g_wi_f,3)`).

Lastly, we can look at how the conventionalized names compared, both between members of the same group and across players in different groups that saw the same color palette of flowers, and whether this differs by condition.

We treat different flowers described by different games as the baseline condition. For individual utility games, descriptions from groupmates are more similar (even for different flowers, `r stats_text(wi_e,8)`), descriptions of the same flower are more similar (even across games, `r stats_text(wi_e,6)`), and there is a large interaction effect, where descriptions of the same flower are very similar among groupmates (`r stats_text(wi_e,7)`).

For shared utility games compared to individual utility games, descriptions are roughly equivalent at baseline (`r stats_text(wi_e,2)`) and for descriptions from groupmates (`r stats_text(wi_e, 5)`). Descriptions are much more similar for the same flower (even across games, `r stats_text(wi_e,3)`), but this is balanced out by a negative interaction effect between same flower and same game (`r stats_text(wi_e,4)`).

These findings are suggestive that in the shared utility condition, games differentiated less from each other, and their conventions may have been influenced more by shared priors for how to describe the flowers.

# Discussion

This analysis of the referring expressions in @mankewitz2021 examined three key questions: the reduction of referring expressions, the convergence of referring expressions to conventions, and the differences between the shared and individual utility conditions.

@@

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found slight condition differences where being in the shared utility condition meant more similar descriptions for the same flowers and more different descriptions for different flowers. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition.

@@

Our interpretation of results is limited by the dataset, which is small, especially compared to recent datasets of convention formation in reference games [@hawkins2020, @boyce2022]. The dataset also has large item-level variation, as the images differ in how nameable the flowers are or whether there are salient features.

This dataset is just the start of testing how well findings from reference games generalize to more complex conversational situations. One dimension that we were unable to explore here due to the design of the game is the idea of shared versus privileged knowledge [@keysar2000taking; @heller2008role]. In the flowers game, like in reference games, all players saw the same set of images (and knew they did), which makes making descriptions to flowers easier. In real-world situations, interlocuters may have different sets of objects in their context and may not know what is in their interlocuters context. An important future direction is studying how repeated reference phenomena interact with different communication forms such as goals and contexts.

In conclusion, this analysis showed that the phenomenon of reduction and conventionalization occurs even in complex games with different incentives, asymmetric knowledge, and more open-ended negotiation and dialogue than that found in reference games. Additionally, we saw hints that different incentives may lightly moderate the formation of conventions, but that reduction and convention formation took place regardless of incentive condition. We hope future work will continue to develop our understanding of these phenomena in more natural settings.

>>>>

---
title: "Communicative reduction in referring expressions within a multi-player negotation game"
bibliography: [ aaflowers.bib]
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

abstract: >
    The ability to form novel conventions is a key signature of efficient linguistic conmmunication. Reduction of referring expressions, one measure of convention formation, is found robustly in dyadic repeated reference games when the target images are initially difficult to name. In reference games, participants share the explicit goal of establishing joint reference. However, establishing reference is a key subgoal of many conversations where interlocuters have more complex goals. In the current work, we analyze a dataset where reference was embedded in strategic 3- player negotiation and coordination games. In these more complex games, we found that the patterns of reduction and convergence to conventions still held across two different incentive conditions, with some suggestive differences between the conditions.

keywords: >
    Convention formation; reference games; reduction of referring expressions; multi-party communication

@@

Successful communication is grounded in a shared understanding of what utterances mean in the context of their production. Whether the communicative goals are cooperative or adversarial, interlocuters need to establish joint reference to be effective. In many cases, conventional word meanings are enough, but there are also contexts where objects to be referred to are not easy to distinguish. In these situations, establishing reference is more difficult, but no less important.

The formation of ad-hoc referring expressions has been studied extensively in the context of dyadic reference games [@clarkReferringCollaborativeProcess1986; @krauss1964changes]. In these, two participants see a set of images (often abstract shapes) and one person is tasked with identifying an image so the other person can pick it out. People can be very successful at this task, achieving high levels of matches. Over repeated reference to the same images, pairs develop shorthand names for the images, leading to shorter descriptions in later repetitions. These nicknames conventionalize with a pair, as they tend to stick to the same description, but are idiosyncratic and differ between pairs. This task has been a well-studied microcosm for understanding reference which has proven useful for testing theories about how referring expressions originate, how expressions are designed, and when they change [@brennan1996conceptual; @metzing2003conceptual; @leung2020parents; @weber2003; @yoonAudienceDesignMultiparty2019].

The implicit theoretical claim is that reduction to partner-specific conventions occurs *whenever* people interact repeatedly in ways that require reference to some objects without adequate conventional names. But do these phenomena generalize beyond artificial reference game environments?

Our current work begins to address this question  by looking for reduction phenomena in a slightly more naturalistic setting. The conversational goals of reference games are not representative of typical language use. In everyday communication, establishing reference is often an instrumental subgoal in a larger conversation. The overall task may be cooperative, such as asking for kitchen equipment while cooking together, or adversarial, such as identifying items while negotiating a division of assets. In contrast, in the reference game context, the goal is explicitly cooperative: match corresponding images.

```{r interface, fig.env = "figure*", fig.pos = "t!", fig.width=6, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Player interface. Panel A shows the selection phase, where each participant sees 6 flowers, 4 with value bars. Panel B shows the feedback stage, after all players have selected. Players see player icons on the value bars indicating who chose what. When multiple players select the same flower, they recieve a lower value rather than what is shown.  \\label{interface}", cache=FALSE}

@@

We build on work by @mankewitz2021 who created a task with the aim of bridging the gap between reference games with hard-to-name objects and negotiation datsets with easily nameable objects. They created a 3-person game where players each selected what flowers to grow from a set of 6 flowers with variable and partially hidden value. Players got the value of the flower only if they were the only one to select it, which incentivized coordination and negotiation. However, all the flowers shown were the same color, so they were not easily nameable, and the flowers repeated, leading to opportunities for ad-hoc conventionalization. This game structure encouraged players to describe flowers while never explicitly telling a player to refer to any specific flower. The experiment had two conditions, one where players within a group won points together (and thus had fully aligned incentives) and another where they won points as individuals (and didn't). @mankewitz2021 found a slight decrease in language over the course of the games, but they did not conduct a detailed analysis to assess whether the decreases in words were associated with shorter and more conventionalized referring expressions. Here, we address this gap by separating out the referring expressions in order to check for patterns of convention-formation.

Our work addresses whether participants in a complex, multi-player strategic game with differing incentives and shared knowledge are able to develop conventionalized reference expressions. We explore whether the phenomena found in classic dyadic iterated reference games extends to this more naturalistic domain, and whether the results differ across the individual and shared incentive conditions. The key questions we addressed were:

@@

Two games consisting of `r raw_rows` utterances were annotated by both annotators.  `r nrow(both)` utterances were identified as containing reference expressions by at least one annotator; of these annotators agreed on the exact reference expression for `r agree_span` (`r round(agree_span/nrow(both)*100)`%) of the cases. The second annotator only coded the target of the referring expression for `r nrow(count)` of the utterances. Of these, the two annotators agreed on the target in `r agree_target` (`r round(agree_target/nrow(count)*100)`%) of cases. We take this level of interannotator agreement as an indication that the reference spans and targets were identified reliably. We extracted a total of `r annotated |> filter(numword>0) |>  tally()` referring expressions.

@@

For descriptions of different flowers, cosine similarity decreased over time (trial: `r stats_text(wi_g_bet_f,6)`, see Figure \ref{during} lower left and middle panels). Utterances being produced by the same person did not have an effect (`r stats_text(wi_g_bet_f,8)`) or an interaction with trial number (`r stats_text(wi_g_bet_f,7)`) .

These results replicate and generalize the pattern found in the prior literature [@hawkins2020].

@@

knitr::kable(text_examples, caption = "Examples of descriptions of different flowers from different games, illustrating reduction and convergence phenomenon. Games are numbered, letters refer to players within games. Images of flowers can be found in Figure 1: flower 1 upper left, flower 2 lower center, flower 3 lower left.\\label{extable}",  format="latex", booktabs=TRUE, linesep="") %>%
kable_styling(full_width = F) %>%
  row_spec(c(3,9), hline_after=T) |> 
column_spec(4, width = "12em") %>%
  kable_styling(latex_options="H")

@@

For descriptions of the same flower across games, we found that cosine similarities decreased over time  (trial: `r stats_text(bet_g_wi_f,4)`, see Figure \ref{during} upper right panel). This decrease is consistent with the predictions.

@@

Referring descriptions later in a game were more similar to the convention than descriptions earlier in the game  (trial: `r stats_text(to_e_wi_f,6)`, see Figure \ref{toend} upper panels). Utterances were more similar to the convention given by the same person (versus the convention given by a groupmate) (`r stats_text(to_e_wi_f,8)`), and this increased over time (`r stats_text(to_e_wi_f,7)`). This is consistent with the expected pattern of convention formation.

@@

### Within games
Games in the shared utilities condition had non-significantly more similar descriptions of the same flower within games (`r stats_text(wi_g_wi_f,2)`), but this did not interact with trial number (`r stats_text(wi_g_wi_f,3)`) or whether the utterances were produced by the same participant (`r stats_text(wi_g_wi_f,5)`). Games in the shared utilities condition did have lower similarities between descriptions of different flowers within games (`r stats_text(wi_g_bet_f,2)`), but this was moderated by trial (`r stats_text(wi_g_bet_f,3)`) and whether the utterances were produced by the same player (`r stats_text(wi_g_bet_f,5)`).

Comparing to the end conventions, games in the shared utility condition produced flower descriptions that were more similar to the convention for that flower (`r stats_text(to_e_wi_f,2)`), but condition did not interact with trial number (`r stats_text(to_e_wi_f,3)`) and did not significantly interact with same participant versus different participant (`r stats_text(to_e_wi_f,5)`).  In the shared utility condition, descriptions were slightly more similar to the conventions for other flowers (`r stats_text(to_e_bet_f,2)`), but condition did not interact with trial (`r stats_text(to_e_bet_f,3)`). This similarity was even higher if the description and convention came from the same person (`r stats_text(to_e_bet_f,5)`). As a whole, groups in the shared utility may have had more internal alignment on the mapping between flowers and names, but this pattern was inconsistent.

### Between games
Across different games, games in the shared utility condition tended to have more similar descriptions for the same flower (`r stats_text(bet_g_wi_f,2)`), and this increased over time (`r stats_text(bet_g_wi_f,3)`).

Lastly, we can look at how the conventionalized names compared, both between members of the same group and across players in different groups that saw the same color palette of flowers, and whether these comparisons differ by condition (Figure \ref{withinend}).

We treated different flowers described by different games as the baseline condition. For individual utility games, descriptions from groupmates were more similar (even for different flowers, `r stats_text(wi_e,8)`), descriptions of the same flower were more similar (even across games, `r stats_text(wi_e,6)`), and there was a large interaction effect, where descriptions of the same flower were very similar among groupmates (`r stats_text(wi_e,7)`).

For shared utility games compared to individual utility games, descriptions were roughly equivalent at baseline (`r stats_text(wi_e,2)`) and for descriptions from groupmates (`r stats_text(wi_e, 5)`). Descriptions were much more similar for the same flower (even across games, `r stats_text(wi_e,3)`), but this was balanced out by a negative interaction effect between same flower and same game (`r stats_text(wi_e,4)`).

These findings are suggestive that in the shared utility condition, games differentiated less from each other, and their conventions may have been influenced more by shared priors for how to describe the flowers than the conventions in individual utility games.

# Discussion
Our main question was whether the reduction phenomena found in dyadic reference games would occur within a negotiation game. This analysis of the referring expressions in @mankewitz2021 examined three key questions: the reduction of referring expressions, the convergence of referring expressions to conventions, and the differences between the shared and individual utility conditions.

@@

Our third consideration was how the different utility conditions moderated the reduction and convergence effects. We found some slight and hard to interpret condition differences within games. In addition, groups in the shared utility condition diverged less from each other. This pattern of condition differences is consistent with less idiosyncratic descriptions in the shared utility condition.

@@

Our interpretation of results is limited by the dataset, which is small, especially compared to recent datasets of convention formation in reference games [@hawkins2020; @boyce2022]. The dataset also has large item-level variation, as the images differ in how nameable the flowers are or how salient background features are.

This dataset is just the start of testing how well findings from reference games generalize to more complex conversational situations. One dimension that we were unable to explore here due to the design of the game is the idea of shared versus privileged knowledge [@keysar2000taking; @heller2008role]. In the flowers game, like in reference games, all players saw the same set of images (and knew they did), which makes making descriptions to flowers easier. In real-world situations, interlocuters may have different sets of objects in their context and may not know what is in their interlocuter's context. An important future direction is studying how repeated reference phenomena interact with different communcation goals and contexts.

In conclusion, our analysis here showed that the phenomenon of reduction and conventionalization occurs even in complex games with different incentives, asymmetric knowledge, and more open-ended negotiation and dialogue than that found in reference games. Additionally, we saw hints that different incentives may lightly moderate the formation of conventions, but that reduction and convention formation took place regardless of incentive condition. We hope future work will continue to develop our understanding of these phenomena in more natural settings.
